---
title: "Final Lab Project"
author: "Lena Nezamzadeh, Ria Bhatia"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    html-table-processing: none
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false 
---

## The Data

In this lab, we will analyze two datasets related to healthcare use in Vietnam at both the individual and household levels. This data comes from the World Bank's Vietnam Living Standards Survey conducted in 1997-1998 and was made available through [Vincent Arel-Bundock ](https://vincentarelbundock.github.io/Rdatasets/datasets.html). 

The larger study can be found [here]( https://microdata.worldbank.org/index.php/catalog/2694).

#### Individual-Level Dataset

Documentation for the individual-level dataset can be found  [here](https://vincentarelbundock.github.io/Rdatasets/doc/camerondata/vietnam_ind.html).


Some variables in this dataset include:

- `injury` – whether the individual experienced an injury during the survey period
- `actdays` – number of days of limited activity
- `pharvis` – number of direct pharmacy visits 
- `lnhhexp` – log of total household expenditure for the year

#### Household-Level Dataset

Documentation for the household-level dataset can be found  [here](https://vincentarelbundock.github.io/Rdatasets/doc/camerondata/vietnam_hh.html).

Some variables in this dataset include:

- `hhsize` – household size 
- `lnhhexp` – log of total household expenditure for the year
- `lnexp12m` – log of household healthcare expenditure over the past 12 months


**0. Load any libraries you think will be useful for this lab.**

```{r setup}
# load libraries

library(tidyverse)
library(knitr)
library(kableExtra)

```

**1. The files for this lab are named `vietnam_hh.csv` and `vietnam_ind.csv`. Load both datasets into R.**

```{r load_data}
# load data

household_data <- read_csv(file = "vietnam_hh.csv")
indiv_data <- read_csv(file = "vietnam_ind.csv")

```

## Cleaning the data

**2a. Individual-Level Data**

**Let’s focus only on the `indiv_data` data. Create a dataset called `clean_indiv_data` using one `dplyr` pipeline that:**

- **renames the `sex` variable to `respondent_male`,**
- **creates a non-log version of household expenditure called `household_exp`,**
- **changes variable types where needed (e.g., `commune` is stored as a numeric variable, but it is really a commune code representing categories rather than numerical values), and**
- **only keeping the columns we will use – `commune`, `respondent_male`, `age`, `educ`, `illness`, `injury`, `illdays`, `insurance`, `lnhhexp`, and `household_exp`.**



```{r}
# solution for question 2a

clean_indiv_data <- indiv_data |> 
  select(commune, age, educ, illness, injury, illdays, 
         insurance, lnhhexp, respondent_male = sex) |> 
  mutate(household_exp = exp(lnhhexp),
         across(.cols = c(commune, respondent_male,
                          injury, insurance),
                .fns = as.factor))
```


**2b. Household-Level Data**

**Now let’s clean the `household_data` data. Create a dataset called `clean_household_data` using one `dplyr` pipeline that:**

- **creates a non-log version of yearly household expenditure called `household_exp`,**
- **creates a non-log version of yearly healthcare expenditure called `healthcare_exp`,**
- **changes variable types where needed (e.g., `commune` is stored as a numeric variable, but it represents categories rather than numerical values), and**
- **only keeps the columns we will use – `commune`, `hhsize`, `lnhhexp`, `lnexp12m`, `household_exp`, and `healthcare_exp`.**

::: {.callout-note}
Some households have missing values for `lnexp12m`. Keep this in mind when calculating summary statistics later in the lab.

Additionally, the original dataset documentation does not clearly specify the units of expenditure after reversing the log transformation. For this lab, interpret `household_exp` and `healthcare_exp` as expenditures on the original (non-log) scale.
:::


```{r}
# solution for question 2b

clean_household_data <- household_data |> 
  select(commune, hhsize, lnhhexp, lnexp12m) |> 
  mutate(household_exp = exp(lnhhexp),
         healthcare_exp = exp(lnexp12m),
         commune = as.factor(commune))

```

## Questions

**3. Using `clean_indiv_data`, calculate the mean number of illness days (`illdays`) for each commune. Arrange the results from highest to lowest average illness days and display only the top 10 communes.**

```{r}
# solution for question 3

clean_indiv_data |> 
  group_by(commune) |> 
  summarise(mean_illdays = mean(illdays)) |> 
  arrange(desc(mean_illdays)) |> 
  slice_head(n = 10) |> 
  kable()
```


**4. Write a function called `commune_spending()` that takes the following inputs:**

- **takes a dataframe,**
- **a commune variable, and**
- **a spending variable**

**The function should calculate the median spending for each commune and return the communes with the highest and lowest median spending.**

::: {.callout-tip}
Hint: Your function returns a list with separate outputs for the highest and lowest median spending communes.
:::

```{r}
# solution for question 4

commune_spending <- function(df, commune_var, spending_var){
  
  spending_summary <- df |>
    group_by({{commune_var}}) |>
    summarise(median_spending = median({{spending_var}}, na.rm = T))
  
  lowest <- spending_summary |>
    slice_min(median_spending, n = 1)

  highest <- spending_summary |>
    slice_max(median_spending, n = 1)
  
  list(lowest = lowest, highest = highest)
}
```

**Run the code below to test your function.**

```{r}

results <- commune_spending(df = clean_household_data,
                 commune_var = commune,
                 spending_var = healthcare_exp)

results$highest
results$lowest
```


**5a. Create a visualization exploring the relationship between average illness days and median household healthcare expenditure across communes.**

**To do this:**

- **calculate the mean number of illness days (`illdays`) for each commune using `clean_indiv_data`,**
- **calculate the median healthcare expenditure (`healthcare_exp`) for each commune using `clean_household_data`,**
- **join the two summary tables together by `commune`,**
- **create a scatterplot showing the relationship between these variables.**

::: {.callout-note}
Your scatterplot should be polished and easy to interpret. Make sure to include informative axis labels and a meaningful title. Since y-axis labels can sometimes be difficult to read vertically, consider whether a subtitle would be a better choice. You may also want to think about adding a smoothing line.
:::

```{r}
# solution for question 5a

commune_illdays <- clean_indiv_data |> 
  group_by(commune) |> 
  summarise(mean_illdays = mean(illdays, na.rm = T))

commune_healthcare <- clean_household_data |> 
  group_by(commune) |> 
  summarise(med_healthcare_exp = median(healthcare_exp, na.rm = T))

commune_illdays |> 
  left_join(commune_healthcare, by = "commune") |> 
  ggplot(aes(x = med_healthcare_exp,
             y = mean_illdays)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm") +
  labs(
    title = "Average Illness Days vs. Median Healthcare Expenditure by Commune",
    subtitle = "Average Illness Days",
    x = "Median Healthcare Expenditure",
    y = ""
  ) 
```


**5b. Across communes, do places with higher healthcare spending also tend to report higher average illness days? Briefly describe the relationship shown in your visualization.**

Solution: Across communes, there appears to be a slight positive relationship between median healthcare expenditure and average illness days. Communes with higher healthcare spending tend to report somewhat higher average illness days, although the relationship is fairly weak and there is still a large amount of variability between communes.


**6. Suppose the probability that an individual has insurance is equal to the proportion observed in `clean_indiv_data`.**

**First calculate the observed insurance rate in the dataset.**

**Then write a function called `simulate_insurance_sample()` that:**

- **takes one input called `sample_size`,**
- **simulates insurance outcomes for that many individuals, and**
- **returns the proportion of simulated individuals who have insurance.**

**Your function should also include input validation to make sure `sample_size` is a single positive number.**

::: {.callout-tip}
Hint: You may find `rbinom()` helpful for simulating binary outcomes like insurance status.
:::

**6a. Below, write the code to calculate observed insurance rate in the data.**

```{r}
# solution for question 6a

prop_ins <- clean_indiv_data |> 
  summarize(obs_insurance_rate = mean(insurance, na.rm = TRUE)) |> 
  pull()

prop_ins
```

**6b. Next, we will write the function. We will use the observed insurance rate as the probability of success in a simulation. In this case, "success" means that a simulated individual has insurance.**

```{r}
# solution for question 6b

simulate_insurance_sample <- function(sample_size){
   if(!is.numeric(sample_size) |
     length(sample_size) != 1 |
     sample_size <= 0){
     stop("Please provide one positive numeric value for sample_size.")}

  simulated_insurance <- rbinom(n = 1, size = sample_size, p = prop_ins)
  
  return(simulated_insurance/sample_size)
}
```

**Run the code below to test your function.**

```{r}
# code to check function works

set.seed(134)
simulate_insurance_sample(100)
```



