---
title: "Final Lab Project"
author: "Lena Nezamzadeh, Ria Bhatia"
format: 
  html:  
    code-tools: true
    toc: true
editor: source
execute: 
  echo: true
  warning: false
  message: false
---

## The Data

In this lab, we’re going to be analyzing two datasets regarding healthcare use in Vietnam at both the individual and household levels. This data was collected from the World Bank’s Vietnam Living Standards Survey of 1997-1998. The data comes to us from [Vincent Arel-Bundock ](https://vincentarelbundock.github.io/Rdatasets/datasets.html). The larger study can be found [here]:( https://microdata.worldbank.org/index.php/catalog/2694)

A detailed distribution of the data at the individual level can be found [here]: (https://vincentarelbundock.github.io/Rdatasets/doc/camerondata/vietnam_ind.html)

These variables include:
- `injury` – 1 if injury during survey period 
- `actdays` – number of days of limited activity
- `pharvis` – number of direct pharmacy visits 
- `lnhhexp` – total household expenditure for the year (ln)

A detailed description of the data at the household level can be found [here]: (https://vincentarelbundock.github.io/Rdatasets/doc/camerondata/vietnam_hh.html)

These variables include: 
- `hhsize` – household size 
- `lnhhexp` – total household expenditure for the year (ln)
- `lnexp12m` – household healthcare expenditure in the past 12 months (ln)


**0. Load the appropriate libraries and the datasets.**

```{r setup}
# load libraries

library(tidyverse)
library(knitr)
library(kableExtra)
library(purrr)

```

```{r load_data}
# load data

household_data <- read_csv(file = "vietnam_hh.csv")
indiv_data <- read_csv(file = "vietnam_ind.csv")

```

## Cleaning the data

**1. Let’s focus only on the `indiv_data` data. Create a dataset called `clean_indiv_data` using one `dplyr` pipeline that:**

- **renames the `sex` variable to `respondent_male`,**
- **creates a non-log version of household expenditure called `household_exp`,**
- **changes data types where needed (e.g., `commune` is stored as a numeric variable, but it is really a commune code representing categories rather than numerical values), and**
- **only keeping the columns we will use – `commune`, `respondent_male`, `age`, `educ`, `illness`, `injury`, `illdays`, `insurance`, `lnhhexp`, and `household_exp`.**



```{r}
# solution 

indiv_data |> 
  select(commune, age, educ, illness, injury, illdays, 
         insurance, lnhhexp, respondent_male = sex) |> 
  mutate(household_exp = exp(lnhhexp)) |>
  mutate(across(.cols = c(commune, respondent_male,
                          injury, insurance),
                .fns = as.factor))
```