---
title: "Final Lab Project"
author: "Lena Nezamzadeh, Ria Bhatia"
format: 
  html:  
    code-tools: true
    toc: true
editor: source
execute: 
  echo: true
  warning: false
  message: false
---

**0. Load the appropriate libraries and the datasets.**

```{r setup}
# load libraries

library(tidyverse)
library(knitr)
library(kableExtra)
library(purrr)

```

```{r load_data}
# load data

household_data <- read_csv(file = "vietnam_hh.csv")
indiv_data <- read_csv(file = "vietnam_ind.csv")

```

## Cleaning the data

**1. Let’s focus only on the `indiv_data` data. Create a dataset called `clean_indiv_data` using one `dplyr` pipeline that:**

- **renames the `sex` variable to `respondent_male`,**
- **creates a non-log version of household expenditure called `household_exp`,**
- **changes data types where needed (e.g., `commune` is stored as a numeric variable, but it is really a commune code representing categories rather than numerical values), and**
- **only keeping the columns we will use – `commune`, `respondent_male`, `age`, `educ`, `illness`, `injury`, `illdays`, `insurance`, `lnhhexp`, and `household_exp`.**



```{r}
# solution 

indiv_data |> 
  select(commune, age, educ, illness, injury, illdays, 
         insurance, lnhhexp, respondent_male = sex) |> 
  mutate(household_exp = exp(lnhhexp)) |>
  mutate(across(.cols = c(commune, respondent_male,
                          injury, insurance),
                .fns = as.factor))
```